---
title: "p8130_hw2"
author: "Chhiring Lama"
date: "2024-10-10"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, message = FALSE}
library(tidyverse)
library(ggplot2)
library(ggthemes)
```

## Problem 1 (15 points)
Suppose the probability of having at least one dental checkup during a two-year period is 73%. 56 random individuals are being studied for their health care utilization practices over a two-year period. Compute “part a” by hand and show key steps. For all other parts you may hand calculate or use R. \

a) What is the probability that exactly 40 of these individuals will have at least one dental
checkup?
- ![]("problem_1a_solution.pdf")\

b) What is the probability that at least 40 of these individuals will have at least one dental
checkup?
```{r}
pdf_until_39 <- pbinom(39, 56, 0.73)
```
- Probability that at least 40 of them have at least 1 check up: $P(X \ge 40) = 1- P(X \le 39)$\
  Using the formula $P(X \le x) = \sum_{x=o}^x {n \choose x}.(p)^x.(1-p)^{n-x}$, we have:\
  $P(X \le 39) = \sum_{x=o}^{39} {56 \choose x}.(p)^x.(1-p)^{56-x}$ = $`r round(pdf_until_39, digits =3)`$\
  
  Hence, the probability of at least 40 of them having at least 1 check up is `r round(pdf_until_39, digits =3)`. \
  
c) Could you use an approximation method to calculate the probabilities above? If yes, calculate the probabilities using approximations and compare to exact values; otherwise, explain why approximation methods are not appropriate.\
- No, we cannot use an approximation method to calculate the probabilities above because number of trials (n) is not large (usually > 100) and probability of success here is quite large (0.73).\

d) How many individuals do you expect to have at least one dental checkup?
```{r expected_val_for_discrete}
prob = function(x){
  pmf = dbinom(x, 56, 0.73)
  x_times_pmf = x*pmf
}
x_val <- seq(from = 0, to = 56)
all_x_times_pmf <- sapply(list(x_val), prob)

expected_val_x <- sum(all_x_times_pmf)
```

- Expected number of individuals to have at least one dental checkup: $E(X)$. Using its formula $E(X) = \sum_{x=0}^n.x.P(X =x)$.\
  In our case, we want to calculate $E(X) = \sum_{x=0}^{56}.x.P(X =x)$. For this we need p.m.f at each x values between 0 and 56. Using R code, $E(X) = `r expected_val_x`$. \
  
e) What is the standard deviation of the number of individuals who will have at least one dental checkup?
```{r expected_val_for_discrete_xsq}
prob_sqx = function(x){
  pmf = dbinom(x, 56, 0.73)
  x_sq_times_pmf = (x^2)*pmf
}

all_x_sq_times_pmf <- sapply(list(x_val), prob_sqx)

expected_val_x_sq <- sum(all_x_sq_times_pmf)

var = expected_val_x_sq - (expected_val_x^2)

sd = sqrt(var)
```
- We know that the $E(x)$ is the population mean($\mu$) as well. To calculate the standard deviation, first we need the variance ($\sigma^2$).\
  $\sigma^2 = E(X^2) - E(X)^2$. From $d)$, we know that $E(X) = `r expected_val_x`$. Using R, we find that $E(X^2) = `r expected_val_x_sq`$, $\sigma^2 = `r round(var, digits = 3)`$ and $\sigma = \sqrt{\sigma^2} = `r round(sd, digits = 3)`$.\
  So, the standard deviation of the number of individuals who will have at least one dental check up is `r round(sd, digits = 3)`.\
  
## Problem 2 (9 points)

Suppose the number of tornadoes in the United States follows a Poisson distribution with parameter $\lambda = 6$ tornadoes per year. Compute using tables or R. Show the formula for "part a" (it can be handwritten and embedded in the pdf file).\

a) What is the probability of having fewer than 3 tornadoes in the United States next year?\
- ![]("problem_2a_solution.pdf")\

b) What is the probability of having exactly 3 tornadoes in the United States next year?\
```{r}
prob_exactly_3 <- dpois(3, 6)
```
- The probability of having exactly 3 tornadoes in the next year is denoted by $P(X = 3)$. Using the formula  $P(X = x) = \frac{\lambda ^x.e^{-\lambda}}{x!}$,\
  $P(X = 3) = \frac{6^3.e^{-6}}{3!} = `r round(prob_exactly_3, digits = 3)`$.\
  The probability of having exactly 3 tornadoes in the next year in the USA is `r round(prob_exactly_3, digits = 3)`. 
  
c) What is the probability of having more than 3 tornadoes in the United States next year?
```{r}
prob_atmost_3 <- ppois(3, 6)
prob_morethan_3 <- 1 - prob_atmost_3
```
- The probability of having more than 3 tornadoes next year $(P(X > 3) = 1 - P(X \le 3))$. \
  $P(X \le 3) = \sum_{x=0}^3\frac{6 ^3.e^{-6}}{3!}$ = `r round(prob_atmost_3, digits = 3)`. \
  $P(X > 3) = 1 - `r round(prob_atmost_3, digits = 3)` = `r round(prob_morethan_3 , digits = 3)`$. \
  Hence, the probability of having more than 3 tornadoes next year in the United States is `r round(prob_morethan_3 , digits = 3)`. 
  
## Problem 3 (12 points)
Assume the systolic blood pressure of 20-29 year old American males is normally distributed with population mean 128.0 and population standard deviation 10.2. \

a) What is the probability that a randomly selected American male between 20 and 29 years old has a systolic blood pressure above 137.0?\
```{r}
x = 137.0
mu = 128.0
sd = 10.2
p_atmost_x <- pnorm(x, mu, sd)
p_lessthan_x <- 1 - p_atmost_x
```
- Here, the population mean systolic blood pressure $(\mu)$ = 128.0 and population standard deviation $(\sigma)$ = 10.2
  P(Systolic\ blood\ pressure > 137.0) = $P(X > 137.0) = 1 - P(X \le 137.0)$ \
  We can calculate c.d.f where for a $X \sim N(128.0, 10.2)$, $f(x) = P(X \le x)$ = $\int_{-\infty}^{x} f(x) dx$ = 
  $\int_{-\infty}^{x} \frac{1}{\sigma\sqrt{2\pi}}.{\frac{-(x-\mu)^{2}}{2\sigma^2}} dx$.\
  
  $P(X \le 137.0)$ = $\int_{-\infty}^{137.0} \frac{1}{10.2\sqrt{2\pi}}.{\frac{-(137.0-128.0)^{2}}{2*10.2^2}} dx$ = $`r round(p_atmost_x, digits = 3)`$. \
  
  So, $P(Z > 137.0)$ = 1 - $P(X \le 137.0)$ = 1 - $`r round(p_atmost_x, digits = 3)`$ = $`r round(p_lessthan_x, digits = 3)`$.\
  The probability that a randomly selected American male between 20 and 29 years old has a systolic blood pressure above 137.0 is $`r round(p_lessthan_x, digits = 3)`$. 

b) What is the probability that the sample mean for blood pressure of 50 males between 20 and 29 years old will be less than 125.0?
```{r}
x_bar = 125.0
n = 50
z_score <- (x_bar-mu)/(sd/sqrt(n))

p_atmost_z <- pnorm(z_score, 0, 1)
```
- Here, we have 50 males (n = 50), and their blodd pressure measure is normally distribute. First, we can get a standardized random variable (Z) corresponding to this mean using $\frac{(\bar X-\mu)}{\frac{\sigma}{\sqrt n}}$, Z = `r round(z_score, digits = 3)`. 
  Using R, we get $P(Z < `r round(z_score,digits = 3)`)$ = `r round(p_atmost_z, digits = 3)`.  
  Therefore, the probability that the sample mean for blood pressure of 50 males between 20 and 29 years old will be less than 125.0 is `r round(p_atmost_z, digits = 3)`. \
  
b) What is the 90th percentile of the sampling distribution of the sample mean X for a sample size of 40?
```{r}
upper_limit = qnorm(0.90, mean = mu, sd = sd)
```

  
  
  


  
